### 2025-08-23 chapters/system-architecture.tex
chapter 3 complete

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.2cm and 1.5cm,
    every node/.style={
        draw,
        rounded corners,
        align=center,
        font=\small,
        minimum width=2.8cm,
        minimum height=1.2cm,
        fill=gray!15,
        drop shadow={shadow xshift=0.5ex,shadow yshift=-0.5ex,opacity=0.1}
    },
    keystone/.style={
        fill=white,
        thick,
        minimum width=3.5cm,
        drop shadow={shadow xshift=0.8ex,shadow yshift=-0.8ex,opacity=0.2}
    },
    connector/.style={
        ->,
        thick,
        >=stealth,
        line width=0.6pt
    }
]

% Nodes
\node (host) [fill=gray!30, minimum height=1.4cm] {Ubuntu 22.04 LTS\\ Host OS};
\node (qemu) [below=of host] {QEMU RV64GC Emulator};
\node (keystone) [keystone, below=of qemu] {Keystone Stack\\ \footnotesize Security Monitor, Runtime, Enclaves};
\node (virtio) [fill=gray!20, left=2cm of keystone] {VirtIO Devices\\ (net, storage)};
\node (plugin) [fill=gray!20, right=2cm of keystone] {Optional Cache Plugin};

% Edges
\draw[connector] (host) -- (qemu);
\draw[connector] (qemu) -- (keystone);
\draw[connector] (virtio) -- (keystone);
\draw[connector] (plugin) -- (keystone);

% Annotation
\node[font=\scriptsize, anchor=south west] at ([xshift=0.3cm,yshift=0.3cm]host.north west) {SSH/GDB access from Host};

\end{tikzpicture}
\caption{Overall virtualized testbed architecture illustrating the host OS, QEMU emulator, Keystone stack, and peripheral components such as VirtIO devices and optional cache plugin.}
\label{fig:virtualized_testbed_architecture}
\end{figure}

The build process utilizes a combination of \texttt{Make} and Buildroot. Buildroot generates the embedded Linux root filesystem, cross-compiles necessary system libraries, and configures the Linux kernel running inside the QEMU guest. The Make-based system orchestrates compilation and deployment of Keystone components such as the runtime, Security Monitor, and kernel modules, ensuring reproducible builds targeting the QEMU RV64 environment.\footnote{Key configuration parameters include the QEMU platform, RV64 CPU architecture, and Buildroot settings optimized for Keystone enclave support.}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.5cm,
    every node/.style={
        draw,
        rounded corners,
        fill=gray!15,
        font=\small,
        minimum width=6cm,
        minimum height=1.2cm,
        align=center
    },
    arrow/.style={
        ->,
        thick,
        >=stealth
    }
]
% Nodes
\node (src) {Source Code \\ Keystone, Kernel, Applications};
\node (buildroot) [below=of src] {Buildroot \\ Rootfs + Kernel};
\node (make) [below=of buildroot] {Make Orchestration \\ Keystone Runtime + Security Monitor};
\node (launch) [below=of make] {QEMU Launch};
\node (bench) [below=of launch] {Run Enclave / Benchmark};

% Arrows
\foreach \from/\to in {src/buildroot, buildroot/make, make/launch, launch/bench} {
    \draw[arrow] (\from) -- (\to);
}
\end{tikzpicture}
\caption{Build and deployment workflow from source code to running enclave and benchmark execution.}
\label{fig:build_deployment_workflow}
\end{figure}

Launching the virtualized system is managed through a Makefile-driven workflow defined in the \texttt{run.mk} file, which encapsulates QEMU configuration and execution commands. Several runtime parameters are exposed as Makefile variables for flexibility, including:

\begin{table}[h]
\centering
\begin{tabular}{l l p{6cm}}
\toprule
\textbf{Parameter} & \textbf{Default} & \textbf{Description} \\
\midrule
QEMU\_PORT & 9821 & SSH port forwarding to guest \\
QEMU\_DBG\_PORT & QEMU\_PORT+1 & TCP port for GDB debugging \\
QEMU\_RAM\_SIZE & 128M & Emulated RAM size \\
QEMU\_CORE\_COUNT & 2 & Number of virtual CPU cores \\
\bottomrule
\end{tabular}
\caption{Key QEMU launch parameters and defaults.}
\end{table}

QEMU is launched with flags specifying the machine type (\texttt{virt}), boot ROM and firmware, kernel and root filesystem images, VirtIO devices for block storage and networking, and network forwarding for SSH access. Optionally, a cache plugin \cite{mandour2021cache} can be enabled to provide detailed logging of cache events such as hits, misses, and evictions. This plugin offers deeper insight into the memory hierarchy’s behavior during enclave execution, aiding analysis of performance impacts due to cache and memory isolation overhead.

The cache plugin \cite{mandour2021cache} parameters are fully configurable at runtime, including:

\begin{lstlisting}[language=bash,caption={Configuration options for cache simulation},label={lst:cache_config}]
# Instruction cache parameters
icachesize=SIZE         # Size of instruction cache
iblksize=BLOCK_SIZE     # Block size
iassoc=ASSOCIATIVITY    # Associativity

@@

For example, to configure the cache plugin with 8 KB L1 instruction and data caches, 64-byte blocks, 4-way associativity, 2 cores, and LRU eviction, the following command is used:

\begin{lstlisting}
make QEMU_PLUGIN_ARGS="dcachesize=8192,dassoc=4,dblksize=64, \
icachesize=8192,iassoc=4,iblksize=64,cores=2,evict=lru" run
\end{lstlisting}

@@

Debugging support is available by enabling the \texttt{KEYSTONE\_DEBUG} flag, which starts QEMU’s built-in GDB server and halts execution at startup for remote debugging. After booting the QEMU guest, the Keystone kernel driver is loaded manually using:

\begin{lstlisting}
modprobe keystone-driver
\end{lstlisting}

@@

Additional Makefile targets enable remote command execution over SSH and connection to the RISC-V GDB debugger for low-level inspection of enclave state. Debugging requires recompiling Keystone with debugging enabled, typically invoked as:

\begin{lstlisting}
KEYSTONE_DEBUG=y make run
make debug-connect
\end{lstlisting}

@@

The Kyber implementation employed in this work is derived from a well-maintained, lightweight C reference library, chosen for its portability and compatibility with embedded systems. This implementation was integrated into the Keystone enclave framework with minimal modification to preserve fidelity to the original cryptographic routines and avoid introducing artifacts that might bias performance measurement. Both the IND-CPA (indistinguishability under chosen-plaintext attack) and IND-CCA (indistinguishability under chosen-ciphertext attack) variants of Kyber were considered, with the primary cryptographic operations — key generation, encapsulation, and decapsulation — being individually invoked inside the enclave to facilitate granular performance profiling.

The following critical Kyber functions were separately instrumented and timed to capture their distinct computational demands:

@@

\begin{lstlisting}[language=bash, caption={CoreMark sample output used to extract IPS}, label={lst:coremark-output}]
2K performance run parameters for coremark.
CoreMark Size    : 666
Total ticks      : 112738928
Iterations/Sec   : 937.43
Iterations       : 100000
\end{lstlisting}

@@

\begin{table}[h]
\centering
\footnotesize
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|l|c|c|c|c|c|c|l|}
\hline
\textbf{Configuration} & \textbf{RAM} & \textbf{Cores} & \textbf{Exec.} &
\textbf{D‑Cache} & \textbf{I‑Cache} & \textbf{Policy} & \textbf{Assoc./Blk (D/I)} \\
\hline
Low‑End Baseline &
64\,MB & 1 & Seq. &
4\,KB & 4\,KB & LRU & 2‑way / 32B,\ 2‑way / 32B \\
\hline
Balanced Parallel &
128\,MB & 2 & Par. &
8\,KB & 8\,KB & FIFO & 4‑way / 64B,\ 4‑way / 64B \\
\hline
Mid‑Range Sequential &
256\,MB & 4 & Seq. &
16\,KB & 8\,KB & Random & 8‑way / 32B,\ 4‑way / 64B \\
\hline
High‑End Parallel &
512\,MB & 6 & Par. &
8\,KB & 16\,KB & LRU & 2‑way / 64B,\ 8‑way / 32B \\
\hline
Max Capacity Stress &
2\,GB & 10 & Par. &
32\,KB & 32\,KB & FIFO & 16‑way / 64B,\ 16‑way / 64B \\
\hline
\end{tabular}
\caption{Summary of representative configurations used in performance evaluation.}
\label{tab:configurations}
\end{table}

@@

In addition to varying memory capacity, core count, and execution mode (sequential or parallel), each configuration was paired with a distinct cache arrangement to assess the sensitivity of enclave and native workloads to cache architecture. The cache parameters are summarised in Table~\ref{tab:configurations} and include:

@@

In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{getrusage()} system call. These included:

>>>>

The experimental procedure followed a systematic pipeline, outlined in Figure~\ref{fig:benchmarking-swimlane}, beginning with benchmark selection and system configuration, followed by dual-mode execution (native and enclave), and concluding with metric collection and comparative analysis.

@@

Figure~\ref{fig:integrated_two_column_peripherals_below} provides a unified representation of the development workflow and the virtualized execution environment used in this thesis. On the left, the build process begins with the Keystone source code and proceeds through root filesystem generation via Buildroot, orchestration of key components through a \texttt{Make}-based system, and culminates in the launch of the full system within a QEMU-emulated RV64GC RISC-V platform. On the right, the runtime stack is depicted in detail: Ubuntu 22.04 LTS acts as the host operating system, supporting QEMU, which emulates the Keystone stack—including the Security Monitor, runtime, and enclave applications.
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    font=\scriptsize,
    node distance=0.8cm and 1.0cm,
    box/.style={
        draw,
        rounded corners,
        fill=gray!15,
        minimum width=3.6cm,
        minimum height=0.9cm,
        align=center
    },
    stack/.style={
        draw,
        rounded corners,
        fill=gray!15,
        minimum width=3.0cm,
        minimum height=0.9cm,
        align=center
    },
    keystone/.style={
        fill=white,
        thick,
        minimum width=3.0cm
    },
    arrow/.style={
        ->,
        thick,
        >=stealth
    }
]

% --- Left column: Build & Deployment Workflow ---
\node (src) [box] {Source Code \\ Keystone, Kernel, Apps};
\node (buildroot) [box, below=of src] {Buildroot \\ Rootfs + Kernel};
\node (make) [box, below=of buildroot] {Make Orchestration};
\node (launch) [box, below=of make] {QEMU Launch};
\node (bench) [box, below=of launch] {Run Enclave / Benchmark};

\foreach \from/\to in {src/buildroot, buildroot/make, make/launch, launch/bench} {
    \draw[arrow] (\from) -- (\to);
}

% --- Right column: Testbed Stack ---
\node (host) [stack, fill=gray!30, minimum height=1.1cm, right=5.0cm of src] {Ubuntu 22.04 LTS\\ Host OS};
\node (qemu) [stack, below=of host] {QEMU RV64GC Emulator};
\node (keystone) [stack, keystone, below=of qemu] {Keystone Stack\\ \footnotesize Sec. Monitor, Runtime, Enclaves};

% Peripherals BELOW Keystone, side-by-side
\node (virtio) [stack, fill=gray!20, below left=0.5cm and 0.5cm of keystone.south] {VirtIO Devices};
\node (plugin) [stack, fill=gray!20, below right=0.5cm and 0.5cm of keystone.south] {Cache Plugin};

% Testbed arrows
\draw[arrow] (host) -- (qemu);
\draw[arrow] (qemu) -- (keystone);
\draw[arrow] (keystone.south) -- ++(0,-0.25) -| (virtio.north);
\draw[arrow] (keystone.south) -- ++(0,-0.25) -| (plugin.north);

% Cross-links from workflow to stack (no intersection)
\draw[arrow, dashed] (launch.east) -- ++(0.8,0) |- (qemu.west);
\draw[arrow, dashed] (bench.east)  -- ++(1.2,0) |- (keystone.west);

% Annotation
\node[font=\tiny, anchor=south west] at ([xshift=0.15cm,yshift=0.15cm]host.north west) {SSH/GDB access};

\end{tikzpicture}
\caption{Overview of the build workflow and virtualized Keystone testbed. The left column shows the step-by-step process of building and launching enclave applications, starting from source code to benchmark execution. The right column depicts the runtime testbed architecture on Ubuntu using QEMU for RISC-V emulation, including the Keystone software stack and supporting peripheral modules (VirtIO and cache plugin). Dashed arrows indicate execution-stage linkage between build steps and their corresponding runtime components.}
\label{fig:integrated_two_column_peripherals_below}
\end{figure}

The design supports modular and reproducible deployment. Peripheral components such as VirtIO devices and an optional cache plugin are attached beneath the enclave runtime, enabling I/O functionality and microarchitectural performance instrumentation. Dashed arrows visually link specific build stages to their corresponding runtime targets, providing a clear and traceable path from compilation to execution. This integration highlights how the build system and execution environment are tightly coupled to support secure enclave development and testing.\footnote{Key configuration parameters include the QEMU platform target, RV64GC architecture, and Buildroot settings customized for Keystone compatibility.}

Launching the system is handled via a Makefile-based interface, which wraps QEMU commands with tunable runtime parameters. The most relevant configuration variables and their default values are summarized in Table~\ref{tab:qemu-params}, including RAM size, core count, and SSH port mapping.
\begin{table}[h]
\centering
\caption{Key QEMU launch parameters and defaults.}
\label{tab:qemu-params}
\begin{tabular}{l l p{6cm}}
\toprule
\textbf{Parameter} & \textbf{Default} & \textbf{Description} \\
\midrule
QEMU\_PORT & 9821 & SSH port forwarding to guest \\
QEMU\_DBG\_PORT & QEMU\_PORT+1 & TCP port for GDB debugging \\
QEMU\_RAM\_SIZE & 128M & Emulated RAM size \\
QEMU\_CORE\_COUNT & 2 & Number of virtual CPU cores \\
\bottomrule
\end{tabular}
\end{table}

Optionally, the cache plugin~\cite{mandour2021cache} can be integrated into the QEMU runtime. This plugin enables logging of cache events—such as hits, misses, and evictions—and is particularly useful when analyzing memory performance of the overall system. Its configuration options are shown in Listing~\ref{lst:cache_config}.
%\vspace{5em}
\begin{lstlisting}[language=bash,caption={Configuration options for cache simulation},label={lst:cache_config}]
# Instruction cache parameters
icachesize=SIZE         # Size of instruction cache
iblksize=BLOCK_SIZE     # Block size
iassoc=ASSOCIATIVITY    # Associativity

@@

For example, Listing~\ref{lst:cache_cmd} demonstrates how to configure an L1 cache setup at runtime.

\begin{lstlisting}[language=bash, caption={Configuration command for cache plugin}, label={lst:cache_cmd}]
make QEMU_PLUGIN_ARGS="dcachesize=8192,dassoc=4,dblksize=64, \
icachesize=8192,iassoc=4,iblksize=64,cores=2,evict=lru" run
\end{lstlisting}

@@

For kernel-level enclave debugging, QEMU’s built-in GDB server can be activated via the \texttt{KEYSTONE\_DEBUG} flag. Once the system boots, the Keystone kernel driver is loaded manually using the command in Listing~\ref{lst:modprobe}.

\begin{lstlisting}[language=bash, caption={Loading the Keystone kernel driver}, label={lst:modprobe}]
modprobe keystone-driver
\end{lstlisting}

@@

In debugging scenarios, the Makefile target \texttt{debug-connect} can be used to attach a GDB session to the running guest. A full debugging session typically starts as shown in Listing~\ref{lst:debug}.

\begin{lstlisting}[language=bash, label={lst:debug}, caption={Starting a QEMU session with GDB debugging enabled}]
KEYSTONE_DEBUG=y make run
make debug-connect
\end{lstlisting}

@@

The Kyber implementation employed in this work is derived from a lightweight, standards-compliant C reference library, selected for its portability and compatibility with embedded systems. This implementation was integrated into the Keystone enclave framework with minimal modification to preserve the correctness and performance characteristics of the original routines, while avoiding structural changes that might distort measurement fidelity.

Both the IND-CPA (indistinguishability under chosen-plaintext attack) and IND-CCA (indistinguishability under chosen-ciphertext attack) security models were considered. For performance profiling, the core cryptographic routines—key generation, encapsulation, and decapsulation—were individually invoked within the enclave. This allowed for fine-grained instrumentation and measurement of each function’s runtime behavior under enclave-induced constraints.

The specific functions evaluated are shown in Listing~\ref{lst:kyber_api}, grouped by security level:

@@

As shown in Listing~\ref{lst:coremark-output}, the CoreMark benchmark outputs the number of iterations per second (IPS), which is a key metric for evaluating system performance.
\begin{lstlisting}[language=bash, caption={CoreMark sample output used to extract IPS}, label={lst:coremark-output}]
2K performance run parameters for coremark.
CoreMark Size    : 666
Total ticks      : 112738928
Iterations/Sec   : 937.43
Iterations       : 100000
\end{lstlisting}

@@

\begin{table}[h]
\centering
\caption{Summary of representative configurations used in performance evaluation.}
\label{tab:configurations}
\footnotesize
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lccccccl}
\toprule
\textbf{Configuration} & \textbf{RAM} & \textbf{Cores} & \textbf{Exec.} &
\textbf{D‑Cache} & \textbf{I‑Cache} & \textbf{Policy} & \textbf{Assoc./Blk (D/I)} \\
\midrule
Low‑End Baseline &
64\,MB & 1 & Seq. &
4\,KB & 4\,KB & LRU & 2‑way / 32B,\ 2‑way / 32B \\
%\hline
Balanced Parallel &
128\,MB & 2 & Par. &
8\,KB & 8\,KB & FIFO & 4‑way / 64B,\ 4‑way / 64B \\
%\hline
Mid‑Range Sequential &
256\,MB & 4 & Seq. &
16\,KB & 8\,KB & Random & 8‑way / 32B,\ 4‑way / 64B \\
%\hline
High‑End Parallel &
512\,MB & 6 & Par. &
8\,KB & 16\,KB & LRU & 2‑way / 64B,\ 8‑way / 32B \\
%\hline
Max Capacity Stress &
2\,GB & 10 & Par. &
32\,KB & 32\,KB & FIFO & 16‑way / 64B,\ 16‑way / 64B \\
\bottomrule
\end{tabular}
\end{table}

@@

In addition to varying memory capacity, core count, and execution mode (sequential or parallel), each configuration was paired with a distinct cache arrangement to assess the sensitivity of enclave and native workloads to cache architecture. The cache parameters are summarised in Table~\ref{tab:configurations} and Listing~\ref{lst:cache_profiles}.

These profiles were designed to represent common cache configurations found in embedded and general-purpose processors, ranging from small, low-associativity caches to larger, highly associative designs. The eviction policies (LRU, FIFO, Random) were selected to explore how different replacement strategies affect performance under varying workload characteristics.

@@

In addition to these high-level metrics, detailed system-level runtime statistics were gathered using standard Linux facilities such as the \texttt{getrusage()} system call.

As shown in Listing~\ref{lst:getrusage}, this call provides insights into CPU time, memory usage, page faults, and context switching events. Collecting these indicators was crucial for diagnosing performance behaviors that could not be explained solely by execution time or throughput.
